---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

## Loading and preprocessing the data

Setting working directory and loading the required libraries.

```{r}
setwd("C:/Data Scientific/Reproducible Research/RepData_PeerAssessment1")
library("data.table")
library("lattice")
```

Reading the .csv using in this research.

```{r}
dt <- read.csv("activity.csv")
```

## What is mean total number of steps taken per day?

Code to calculate the total number of steps for each day.

```{r}
#this function give the sum of each interval in one day
totalPerDay <- function(date)
{
  sum(dt$steps[dt$date == date], na.rm = TRUE)
}
#using a sapply to generate the total of steps for each day
means <-  data.table(date=as.POSIXct(unique(dt$date)),"stepsMean"=sapply(unique(dt$date),totalPerDay))
```

Histogram with the mean steps taken each day.

```{r, scatterplot, echo=FALSE} 
histogram(means$stepsMean,xlab = "Total steps for each day", type="count")
```

```{r, echo=FALSE}
showMean <- mean(means$stepsMean)
showMedian <- median(means$stepsMean)
```

The median of steps for eachs day is `r showMedian`.
The average of steps is `r showMean`.

## What is the average daily activity pattern?

```{r, echo=FALSE}
#this chunk code provides differents measures for each day
#note the days are in spanish because my R is in spanish
#
#this code chunk is unused because I choose other strategy more easy to
#solve the problem
means$weekday <- weekdays(means$date)

averagePerDay <- data.table (day = unique(weekdays(as.POSIXct(dt$date))))

meanDay <- NULL

s1 <- as.list(subset(means,weekday == "lunes",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "martes",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "miércoles",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "jueves",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "viernes",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "sábado",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

s1 <- as.list(subset(means,weekday == "domingo",select = stepsMean))
meanDay <- c(meanDay,mean(s1$stepsMean, na.rm = TRUE))

averagePerDay$mean <- meanDay
```

This code chunk calculate the average of the same interval over all the days.

```{r}
totalPerInterval <- function(interval)
{
  mean(dt$steps[dt$interval == interval], na.rm = TRUE)
}
activity <-  data.table(interval=unique(dt$interval),"stepsMean"=sapply(unique(dt$interval),totalPerInterval))
```

Time-series plot with the average activity pattern 

```{r, echo=FALSE}
plot(activity$interval,activity$steps,type='l', xlab="interval", ylab="steps",main = "Average activity pattern")
```

```{r, echo=FALSE}
interval_with_max_steps <- activity[activity$stepsMean == max(activity$stepsMean),]
```

The interval with the average maximun number of steps  is the interval number `r interval_with_max_steps$interval` with  `r interval_with_max_steps$stepsMean` steps

## Imputing missing values

Number total of values with NA using is.na.data.frame

```{r}
numberOfNAS <- is.na.data.frame(dt)
sum(numberOfNAS)
```

Strategy for filling the missing values of the dataset:

- Replace the missing values with the average of steps per day between the number of intervals

Create a new dataset called newDt with the NAS filled.
```{r}
value_to_fill <- showMean / length(unique(dt$interval))
newDt <- dt
newDt$steps[is.na(newDt$steps)] <- value_to_fill 
```

Check again the amount of NAS.
```{r}
numberOfNAS <- is.na.data.frame(newDt)
sum(numberOfNAS)
```

Code to calculate the total number of steps for each day.

```{r}
#this function give the sum of each interval in one day
totalPerDay2 <- function(date)
{
  sum(newDt$steps[newDt$date == date], na.rm = TRUE)
}
#using a sapply to generate the total of steps for each day
meansNew <-  data.table(date=as.POSIXct(unique(newDt$date)),"stepsMean"=sapply(unique(newDt$date),totalPerDay2))
```

Histogram with the mean steps taken each day without NAs.

```{r, scatterplot2, echo=FALSE} 
histogram(meansNew$stepsMean,xlab = "Total steps for each day", type="count")
```

```{r, echo=FALSE}
showMean2 <- mean(meansNew$stepsMean)
showMedian2 <- median(meansNew$stepsMean)
```

The median of steps for eachs day is now `r showMedian2` with NAs is `r showMedian`.
The average of steps is now `r showMean2` with Nas is `r showMean`.

We can observe the median still being the same and the average is raised.

## Are there differences in activity patterns between weekdays and weekends?

With the new dataset we added the kindOfDay factor value.

This code chunk added a factor of 'weekend' or 'weekday' in a field called kindOfDay.

```{r}
fillKindOfDay <- function(dateIn)
{
    day = weekdays(as.POSIXct(dateIn))
    if (day == "lunes" || day == "martes" || day == "miércoles" || day == "jueves" || day == "viernes")
    {
      return("weekday")
    }
    else if (day == "Monday" || day == "Tuesday" || day == "Wednesday" || day == "Thursday" || day == "Friday")
    {
      return("weekday")
    }
    else if (day == "sábado" || day == "domingo")
    {
      return("weekend")
    }
    else if (day == "Saturday" || day == "Sunday")
    {
      return("weekend")
    }
    else
    {
      return(NA)   
    }    
}
newDt$kindOfDay <-  sapply(newDt$date,fillKindOfDay)
head(newDt)
```

```{r, echo=FALSE}
totalPerIntervalWeekday <- function(interval)
{
  mean(newDt$steps[newDt$interval == interval & newDt$kindOfDay == "weekday"], na.rm = TRUE)
}
totalPerIntervalWeekend <- function(interval)
{
  mean(newDt$steps[newDt$interval == interval & newDt$kindOfDay == "weekend"], na.rm = TRUE)
}
```

Plot with the average activity patterns.

```{r, echo=FALSE}
par(mfcol = c(1,1))
activity <-  data.table(interval=unique(dt$interval),"stepsMean"=sapply(unique(dt$interval),totalPerIntervalWeekday))
plot(activity$interval,activity$steps,type='l', xlab="interval", ylab="steps",main = "Average activity pattern (Weekday)")
activity <-  data.table(interval=unique(dt$interval),"stepsMean"=sapply(unique(dt$interval),totalPerIntervalWeekend))
plot(activity$interval,activity$steps,type='l', xlab="interval", ylab="steps",main = "Average activity pattern (Weekend)")
```
